---
title: "ESM 244 NLS Lab"
author: "Nathaniel Grimes, Casey O'Hara"
date: "1/19/2023"
output: html_document
---

```{r setup, include=TRUE, message= FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning = FALSE)
library(purrr)
library(tidyverse)
library(Metrics)
library(cowplot)
library(here)


```


```{r}
load(here("labs","data","fish_data.Rdata"))
```


## Single model NLS

\begin{equation}
S_t=B_{t+1}-B_t+C_t
\end{equation}


```{r}
surplus<-Fish_data %>% 
  group_by(stockid) %>% 
  select(stockid,year,TBbest,TCbest) %>% 
  drop_na() %>% 
  mutate(f_biomass=lead(TBbest)) %>% 
  mutate(surplus=f_biomass-TBbest+TCbest) %>% 
  drop_na()
  
  
```


Let's see what our data looks like with an example of one stock.

```{r}
one_stock<-surplus %>% 
  filter(stockid=="COD1f-XIV")

ggplot(data=one_stock,aes(x=year,y=surplus))+
  geom_point(size=3,color="black")+
  theme_minimal()

```

### Step One: Selecting a Model

A Fox model
\begin{equation}
\hat{S_t}=-e*MSY(\frac{B_t}{K})\ln(\frac{B_t}{K})
\end{equation}

Where e is base of the natural log $\approx$ 2.718, MSY is the maximum sustainable yield, K is the carrying capacity, and $B_t$ is the biomass for the observed year.


```{r foxmodel}
fox<-function(m,carry,biomass){
 out= -2.718*m*(biomass/carry)*log(biomass/carry)
return(out)
}
```


### Step Two: Initial Guess


```{r guess}

#Write out the guess first, we'll move into the nls wrapper soon

#Since MSY comes into the fox function first, it needs to be the first guess

guess_vec=c(max(one_stock$TBbest)*0.37,
            max(one_stock$TBbest))

```

### Step Three: Run NLS


```{r nlsonemodel}
one_stock_nls=nls(surplus~fox(m,carry,TBbest),
                  data=one_stock,
                  start=list(m=guess_vec[1],carry=guess_vec[2]),
                  trace=TRUE)
```

Show how the model fits the data.

```{r}
# Make a predict based on the model

one_stock_predict<-one_stock %>% 
  mutate(predict=predict(one_stock_nls,newdata=.))

ggplot(data=one_stock_predict)+
  geom_point(aes(x=year,y=surplus))+
  geom_path(aes(x=year,y=predict),color='red')
```


## Using purrr to run many nls models


```{r nlsmany}
#Define a new function to pass along the nls calls

all_nls_fcn<-function(surplus_df){
  nls(surplus~fox(m,carry,TBbest),
  data=surplus_df,
  start=list(m=max(surplus_df$TBbest)*0.37,carry=max(surplus_df$TBbest)))
}

## Pay attention to the position and use of .x, .y, and .f in the map functions

# stop and show the class how each pipe changes the dataframe after nest()

fox_all<-surplus %>%
  group_by(stockid) %>% 
  nest() %>% 
  mutate(nls_model=map(data,~all_nls_fcn(.x))) %>% 
  mutate(predictions=map2(nls_model,data,~predict(.x,newdata=.y))) %>% 
  mutate(rmse=map2_dbl(predictions,data,~rmse(.x,.y$surplus)))

```

## Compare to a random null model


```{r}

# Define the model, don't worry to much how I got it what it means
r_avg<-function(surplus){
  avg_sur=mean(surplus)
  
  rmse=sqrt(mean((avg_sur-surplus)^2))
  
  return(rmse)
}


r_mse<-surplus %>%
  group_by(stockid) %>% 
  nest() %>% 
  mutate(rmse=map_dbl(data,~r_avg(.x$surplus)))
```

## How did the models compare to the null?

```{r}
which(r_mse$rmse-fox_all$rmse<0)

fox_all$stockid[39]
```

## Graph the top 5 best fit models



```{r}
plots<-fox_all %>% 
  arrange(rmse) %>% 
  head(5) %>% 
  mutate(graph=map2(data,predictions,
                    ~ggplot()+
                      geom_point(data = .x,aes(x=.x$year,y=.x$surplus,color='Actual'))+
                      geom_point(aes(x=.x$year,y=.y,color='Predicted'))+
                      theme_minimal()+xlab('')+ylab('Surplus')+
                      scale_color_manual(name="Legend",breaks = c('Actual','Predicted'),values=c('Actual'='black','Predicted'='red'))))

#Make a new list
plot_list=plots$graph

cowplot::plot_grid(plotlist=plot_list,labels =c( plots$stockid,""),hjust=-0.5,vjust=1)


```

### Extra Cowplot fun to make it look really clean

```{r legends}
#extract the legend

legend<-get_legend(plots$graph[[1]])

#remove the legend from the plots
# sometimes you don't always need to use purrr, a quick for loop for something easy can also work
for(i in 1:length(plots$graph)){
  plots$graph[[i]]<-plots$graph[[i]]+theme(legend.position = "none")
}

#Make a new list
plot_list_legend=plots$graph

#create a new plot in the empty space of cowplot grids and fill it with the legend info we took from the plots.

plot_list_legend[[6]]<-legend

cowplot::plot_grid(plotlist=plot_list_legend,labels =c( plots$stockid,""),hjust=-0.5,vjust=1)
```

