---
title: "Assignment 2 Task 2 Answer Key"
author: "Nathan Grimes"
date: "1/24/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE,eval=TRUE)

library(tidyverse)
library(cowplot)
library(Metrics)
library(cowplot)
library(modelr)
```

## Task 2: Parameter Estimation with Purrr - Lizards Length to Weight

Source: Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f

For task 2, you will use non linear least squares to estimate parameters of a length to weight model for lizard populations in New Mexico. 


Create a knitted.html saved as `a2_task2_lastname_firstname.html` which provides a description of the dataset, the purpose of the analysis, a citation of the dataset, and answers the following questions.

###

1) Create a dataframe of only the sex, species, snout to vent length, and weight. Exclude juveniles from the analysis. Remove any species that possess less than 6 observations. Finally, make sure you "balance" the data between male and female observations (e.g. make sure all species that are selected have observations of both male and female). Show the list of the species that will be used in the rest of the task (can just print out, no need for a table)

1) Fit a snout length to weight model of the following form to all lizard in your clean dataframe.

\begin{equation}
W=a(SVL)^b
\end{equation}

  a) Weight is given by W, snout to vent length by SVL, and a and b are the parameters that need to be fitted. Which strategy would be best to provide an initial guess? We could go with strategy one and look through the literature, but let’s practice our coding and math skills.
  
  b) Since we know the model is exponential in nature, we could log transform the data. If we do a standard OLS regression on the log transformed data, we can get approximations of the parameters from the regression coefficients
`my_guess_model <- lm(log_weight ~ log_length, data = my_df) `

  c) Using the coefficients function, we can then supply the nls start list with the regression coefficients. Hint: You will have to mathematically transform the intercept coefficient to get the guess for parameter a. 

```{r part1}
library(tidyverse)

# Here is the wrangling I did to get the data in a usable form

lizard <- read_csv(here::here("assignments","assignment-2","lizard.csv")) #Note this data comes directly from LTER

#Remove juveniles, na, and observations less than 6
lizard_clean<-lizard %>% 
  select(spp,sex,SV_length,weight) %>% 
  filter(sex!="J") %>% 
  drop_na() %>% 
  group_by(spp,sex) %>% 
  filter(n()>=6)

#get names of species that have both male and female for balanced analsyis

lizard_m<-lizard_clean %>% 
  filter(sex=="M")

lizard_f<-lizard_clean %>% 
  filter(sex=="F")

spp_select<-intersect(unique(lizard_m$spp),unique((lizard_f$spp)))
print(spp_select)


# create final combined df for model use in purrr and nls
lizard_comb<-lizard_clean %>%   #Lizard comb is what the students recieved
  filter(spp %in% spp_select)
```


2) Present your fitted model on a plot with female and male lizards separated by color. You should include the nls model in `kable` output of the html. 

```{r part2}
#Create length function model
length_to_weight<-function(a,b,length){
  out=a*length^b
  
  return(out)
}
# Add controls
controls=nls.control(maxiter=100)

guess_model=lm(log(weight)~log(SV_length),data=lizard_comb)

#Run nls
nls_one<-nls(weight~length_to_weight(a,b,SV_length),
             data=lizard_comb,
             start=list(a=exp(coefficients(guess_model))[1],b=coefficients(guess_model)[2]),
             control=controls)

#Set up the predictions for graphing
length_step<- seq(from=min(lizard_comb$SV_length),to=max(lizard_comb$SV_length),length.out=100)

predictions<-length_to_weight(a=coefficients(nls_one)[1],b=coefficients(nls_one)[2],length_step)

#collect the predictions into one df
dfplot=data.frame(predict=predictions,length=length_step)
  
#Make the graph, but make it look pretty than this
ggplot(data=lizard_comb)+
  geom_point(aes(x=SV_length,y=weight,color=sex))+
  geom_line(data=dfplot,aes(x=length,y=predict),size=2.5)+
  theme_minimal()


```

3) Filter out the dataset for male “CNTI” species. Fit an nls model to this subset. Compare the output from the species specific nls model to the general nls model for all species by graphing the model fits on the “CNTI” male data. Report the RMSE for both models in the figure caption. Also in the figure caption briefly discuss which model should be used and why.

```{r compare}

best<-lizard_comb %>% 
  filter(spp=="CNTI"&sex=="M")

guess_model=lm(log(weight)~log(SV_length),data=best)

#Run nls
nls_best<-nls(weight~length_to_weight(a,b,SV_length),
             data=best,
             start=list(a=exp(coefficients(guess_model))[1],b=coefficients(guess_model)[2]),
             control=controls)

length_step<- seq(from=min(best$SV_length),to=max(best$SV_length),length.out=100)

predictions_comb<-length_to_weight(a=coefficients(nls_one)[1],b=coefficients(nls_one)[2],length_step)

predictions_best<-length_to_weight(a=coefficients(nls_best)[1],b=coefficients(nls_best)[2],length_step)

#collect the predictions into one df
dfplot=data.frame(predict=predictions_comb,length=length_step,best=predictions_best)


## Try with modelr

pred<-best %>% gather_predictions(nls_best,nls_one) 

ggplot(data=pred)+
  geom_point(aes(x=SV_length,y=weight),color="black",size=3)+
  geom_line(aes(x=SV_length,y=pred,color=model),size=3)+
  scale_color_manual(values=c("blue","red"))+
  theme_minimal()



### find RMSE between the two models

predict_best<-predict(nls_best,newdata = best$SV_length)
predict_comb<-predict(nls_one,newdata=best$SV_length)

rmse_best<-signif(Metrics::rmse(best$weight,predict_best))
rmse_comb<-signif(Metrics::rmse(best$weight,predict_comb))

rmse_mod<-signif(rmse(nls_best,best))
rmse_mod2<-signif(rmse(nls_one,best))  ## From modelr

```

Using the Metrics package I get RMSE of `r rmse_best` for single species model and `r rmse_comb` for the all species. With modelr the single species model is nearly identical with `r rmse_mod` to the all species nls rmse of `r rmse_mod2`.


### Optional 

4) Group by species and sex then nest the data. Use `map()` variants to parameterize length to weight models for every species separated by each sex. Hint: you may need to adjust the controls of the `nls` function

```{r part3}
#make nls function for purrr

nls_fcn<-function(lizard_df){
  
  guess_model=lm(log(weight)~log(SV_length),data=lizard_df)
  
  nls_one<-nls(weight~length_to_weight(a,b,SV_length),
               data=lizard_df,
               start=list(a=exp(coefficients(guess_model)[1]),b=coefficients(guess_model)[2]),control = controls)
  
  return(nls_one)
}

nls_lizards_purr<-lizard_comb %>% 
  group_by(spp,sex) %>% 
  nest() %>% 
  arrange(spp) %>% 
  mutate(nls_model=map(data,~nls_fcn(.x))) %>% 
  mutate(predictions=map2(nls_model,data,~predict(.x,newdata=.y))) %>%  #Part 4 here
  mutate(RMSE=map2_dbl(predictions,data,~Metrics::rmse(.x,.y$weight))) %>% 
  mutate(coeff=map(nls_model,~coefficients(.x)))   #helps with part 6
```


```{r}
nls_lizards_purr<-lizard_comb %>% 
  group_by(spp,sex) %>% 
  nest() %>% 
  arrange(spp) %>% 
  mutate(nls_model=map(data,~nls_fcn(.x))) %>% 
  mutate(predictions=map2(nls_model,data,~predict(.x,newdata=.y))) %>%  #Part 4 here
  mutate(RMSE=map2_dbl(nls_model,data,~rmse(.x,.y))) %>% 
  mutate(mape=map2_dbl(nls_model,data,~mape(.x,.y))) %>% 
  mutate(coeff=map(nls_model,~coefficients(.x))) %>% 
  mutate(one_rmse=map_dbl(data,~rmse(nls_one,.x))) %>% 
  mutate(one_mape=map_dbl(data,~mape(nls_one,.x))) %>% 
  mutate(one_rsq=map_dbl(data,~rsquare(nls_one,.x))) %>% 
  mutate(rsq=map2_dbl(nls_model,data,~rsquare(.x,.y))) %>% 
  mutate(pred_one=map(data,~predict(nls_one,newdata=.x)))
```


5) Use the models to predict the data. Calculate the RMSE for each model created in part 3. Make a table showing the RMSE for each model.

6) Create a plot grid showing each species and sexes actual weight and the model prediction.

```{r part5}
  plts<-nls_lizards_purr %>%
  mutate(graph=map2(data,predictions,~ggplot()+
                geom_point(data = .x,aes(x=SV_length,y=weight))+
                geom_line(aes(x=.x$SV_length,y=.y),color="red",size=2)+
                  labs(y="Weight",x="Snout Length\n")+
                  theme_half_open()))



labels=nls_lizards_purr %>% 
  mutate(long_label=case_when(sex=="F"~"Female",
                              sex=="M"~"Male")) %>% 
  mutate(labels=paste(spp,long_label,sep = " ")) %>% 
  ungroup() %>% 
  select(labels)
  
  cowplot::plot_grid(plotlist = plts$graph,labels = labels$labels,label_size=10,vjust=0.8)
  
  #patchwork::wrap_plots(plts$graph)
```

```{r}
  plts_2<-nls_lizards_purr %>% 
    select(data,predictions,pred_one) %>% 
    rename(liz=data) %>% 
    unnest_longer(c(liz,predictions,pred_one)) %>% 
    nest() %>% 
  mutate(graph=map(data,~ggplot(data=.x)+
                geom_point(aes(x=liz$SV_length,y=liz$weight))+
                geom_line(aes(x=liz$SV_length,y=predictions,color="Single Species Model"),size=2)+
                geom_line(aes(x=liz$SV_length,y=pred_one,color="All Species Model"),size=2)+
                  scale_color_manual(name="",values=c("blue","red"))+
                  labs(y="Weight",x="Snout Length\n")+
                  theme_half_open()))

legend<-get_legend(plts_2$graph[[1]])

#remove the legend from the plots
# sometimes you don't always need to use purrr, a quick for loop for something easy can also work
for(i in 1:length(plts_2$graph)){
  plts_2$graph[[i]]<-plts_2$graph[[i]]+theme(legend.position = "none")
}

#Make a new list
plot_list_legend=plts_2$graph

#create a new plot in the empty space of cowplot grids and fill it with the legend info we took from the plots.

plot_list_legend[[16]]<-legend

cowplot::plot_grid(plotlist=plot_list_legend,labels = labels$labels,vjust=1,label_size=10)


```


7) Calculate the percent difference in the $b$ parameter between male and females. Plot the differences as a graph. Describe the implications of your result.

```{r part6}
coeff<-nls_lizards_purr %>%
  select(spp,sex,coeff) %>% 
  unnest(coeff) %>% 
  group_by(spp,sex) %>% 
  arrange(sex) %>% 
  mutate(female_b=lead(coeff)) %>% 
  drop_na() %>% 
  group_by(spp) %>% 
  mutate(male_a=lead(coeff)) %>% 
  mutate(male_b=lead(female_b)) %>% 
  drop_na() %>% 
  mutate(pctdiff_a=(male_a-coeff)/coeff) %>% 
  mutate(pctdiff_b=(male_b-female_b)/female_b)


ggplot(data=coeff,aes(x=spp,y=pctdiff_b))+
  geom_col()+
  scale_y_continuous(labels=scales::percent)+
  labs(y="Percent Difference in b parameter",x="Lizard Species")+
  theme_minimal()

 

  
```


Comparing the two graphs the possible importance, but also the strength of nls. In the datasets with limited data, the model predictions are the most different as well as the parameters (see PHMO species). The species I choose by chance is nearly identical in parameter values between males and females as well as being almost perfectly correlated with the total nls model. I was trying to demonstrate that if we have enough information on the species it might be best to run a nls just on that species. If we have a new species though, the model is still pretty predictive so it depends on your confidence in identifying lizard species.


